---
title: "Hillshade Effects (orig. by Dominic Royé)"
author: "Antti Rask"
date: "2023-07-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Hillshade effects

This R Markdown document is me applying the code in __Dominic Royé__'s [blog post](https://dominicroye.github.io/en/2022/hillshade-effects/) about the _hillshade_ shadow effects on relief maps. The twist is that the original maps Dominic uses are of Switzerland and I chose to use the map of Finland, since that's where I hail from.

The three big changes that came from the country change were as follows:

1. Had to search for the Finnish lake boundaries and found the data on the Finnish Environmental Institute [website](https://www.syke.fi/en-US/Open_information/Spatial_datasets/Downloadable_spatial_dataset).
2. Due to the different shapes that Switzerland and Finland have, I decided to move the legend to the right side of the map instead of the bottom.
3. And despite the lyrics in the popular song _Finland, Finland, Finland_ by __Monty Python__, Finland doesn't really have any _'lofty'_ mountains, so had to also adjust the elevations in the legend.

With that, let's get going. I have to warn you that I will keep the commenting to a minimum (mainly to tell you where I've made changes to the original code). If you wish to better understand what's going on at each stage, I suggest you check out Dominic's original [blog post](https://dominicroye.github.io/en/2022/hillshade-effects/).

## Packages

```{r, message=FALSE, warning=FALSE}
library(conflicted) # An Alternative Conflict Resolution Strategy
conflicts_prefer(dplyr::filter)
library(elevatr)    # Access Elevation Data from Various APIs
library(ggblend)    # Blending and Compositing Algebra for 'ggplot2'
library(ggnewscale) # Multiple Fill and Colour Scales in 'ggplot2'
library(giscoR)     # Download Map Data from GISCO API - Eurostat
library(sf)         # Simple Features for R
library(terra)      # Spatial Data Analysis
library(tidyterra)  # 'tidyverse' Methods and 'ggplot2' Helpers for 'terra' Objects
library(tidyverse)  # Easily Install and Load the 'Tidyverse'
library(units)      # Measurement Units for R Vectors
library(whitebox)   # 'WhiteboxTools' R Frontend
```

## Data

```{r}
fin  <- gisco_get_countries(
    country    = "Finland", 
    resolution = "03" # 1:3 million
)

plot(
    fin,
    max.plot = 1 # otherwise R will print one map for each column
)
```

As I mentioned in the intro, I found the data for the Finnish lake boundaries on the Finnish Environmental Institute [website](https://www.syke.fi/en-US/Open_information/Spatial_datasets/Downloadable_spatial_dataset)).

The file is _Jarvi10.shp_ and is included in the __Shoreline - Lakes (Ranta10)__ [.zip file](https://wwwd3.ymparisto.fi/d3/gis_data/spesific/ranta10jarvet.zip).

```{r}
# import the lakes boundaries
lakes  <- st_read("data/Jarvi10.shp")

lakes
```

Finland has more lakes than Switzerland, but I ended up using the same filtering principle of lake area being over 50 km2. _pAla_ha_ is the column with the area of the lakes, but it's in hectares, so first we need some conversion.

Likewise, I'm only using lakes that are inside the Finnish borders. _valtio_ translates to country code and we can use it for that.

```{r}
## filter the largest ones ----
lakes <- lakes %>%
    mutate(areakm = set_units(pAla_ha, "hectare") %>% set_units("km2")) %>% 
    arrange(desc(areakm)) %>%
    filter(
        valtio == "FI",
        areakm > set_units(50, "km2")
    )

lakes
```

```{r}
plot(
    lakes,
    max.plot = 1
)
```

## Digital Elevation Model (DEM)

```{r}
# get the DEM with get_elev_raster
mdt <- get_elev_raster(
    fin,
    z = 8 # the bigger the z (zoom), the longer it takes to load
)

mdt
```

```{r}
plot(mdt)
```

```{r}
# convert to terra and mask area of interest
mdt <- mdt %>%
    rast() %>%
    mask(vect(fin))

# reproject
mdt <- mdt %>% 
    project(crs(lakes))

# reproject vect
fin <- fin %>% 
    st_transform(st_crs(lakes))

fin
```

```{r}
# convert the raster into xyz
mdt_tbl <- as.data.frame(
    mdt,
    xy = TRUE
) %>% 
    as_tibble()

names(mdt_tbl)[3] <- "alt"

mdt_tbl
```

We have to change the scale for the legend, since, as mentioned before, Finland doesn't have any mountains.

Also, since we're already using the _guides()_ function to manipulate the different aspects of the legend, we can double down on that and leave out the _themes()_ function altogether.

And as mentioned before, the legend is vertical, due to the shape of Finland, compared to Switzerland.

All of the above is true for each of the maps, but I won't repeat myself before each of them. Now you know.

```{r}
# map
p1 <- mdt_tbl %>%
    ggplot() +
    geom_raster(aes(x, y, fill = alt)) +
    geom_sf(
        data  = lakes,
        fill  = "#c6dbef",
        color = NA
    ) +
    scale_fill_hypso_tint_c(
        breaks = c(
            100, 200, 300, 400, 500, 600, 700, 800, 900, 1000
        )
    ) +
    guides(
        fill = guide_colorsteps(
            barwidth       = .5,
            barheight      = 20,
            title          = "m",
            title.position = "top",
            direction      = "vertical"
        )
    ) +
    coord_sf() +
    theme_void()

p1
```

The .png files in the __img__ folders have been made with the _ggsave()_ function. I've commented out the code. Unless you make changes to the maps themselves, there's no need to run any of them.

Again, due to having a different shape maps, I decided to use the vertical setup (A4) for the .png files. 

```{r}
# ggsave(
#     "img/01_finland_DEM.png",
#     p1,
#     width  = 8.3,
#     height = 11.7,
#     unit   = "in",
#     device = png,
#     type   = "cairo",
#     bg     = "white"
# )
```

## Calculate the hillshade

```{r}
## estimate the slope ----
sl <- terrain(
    mdt,
    "slope",
    unit = "radians"
)

plot(sl)
```

```{r}
## estimate the aspect or orientation ----
asp <- terrain(
    mdt,
    "aspect",
    unit = "radians"
)

plot(asp)
```

```{r}
# calculate the hillshade effect with 45º of elevation
hill_single <- shade(
    sl,
    asp,
    angle     = 45,
    direction = 300,
    normalize = TRUE
)

# final hillshade
plot(
    hill_single,
    col = grey(1:100/100)
)
```

## Combine the relief and shadow effect

```{r}
# convert the hillshade to xyz
hill_single_tbl <- as.data.frame(
    hill_single,
    xy = TRUE
) %>% 
    as_tibble()

hill_single_tbl
```

```{r}
# map
p2 <- hill_single_tbl %>%
    ggplot() +
    geom_raster(
        aes(x, y, fill = hillshade),
        show.legend = FALSE
    ) +
    scale_fill_distiller(palette = "Greys") +
    new_scale_fill() +
    geom_raster(
        data = mdt_tbl,
        aes(x, y, fill = alt),
        alpha = .7
    ) +
    scale_fill_hypso_tint_c(
        breaks = c(
            100, 200, 300, 400, 500, 600, 700, 800, 900, 1000
        )
    ) +
    geom_sf(
        data = lakes,
        fill = "#c6dbef",
        color = NA
    ) +
    guides(
        fill = guide_colorsteps(
            barwidth       = .5,
            barheight      = 20,
            title          = "m",
            title.position = "top",
            direction      = "vertical"
        )
    ) +
    coord_sf() +
    theme_void()

p2
```

```{r}
# ggsave(
#     "img/02_finland_relief_and_shadow.png",
#     p2,
#     width  = 8.3,
#     height = 11.7,
#     unit   = "in",
#     device = png,
#     type   = "cairo",
#     bg     = "white"
# )
```

## Multidirectional shadows

```{r}
# pass multiple directions to shade()
hill_multi <- map(c(270, 15, 60, 330), function(dir){ 
    
    shade(
        sl,
        asp, 
        angle     = 45, 
        direction = dir,
        normalize = TRUE
    )
})

# create a multidimensional raster and reduce it by summing up
hill_multi <- rast(hill_multi) %>%
    sum()

# multidirectional
plot(
    hill_multi,
    col = grey(1:100/100)
)
```

```{r}
# convert the hillshade to xyz
hill_multi_tbl <- as.data.frame(
    hill_multi,
    xy = TRUE
) %>%
    as_tibble()

hill_multi_tbl
```

```{r}
# map
p3 <- ggplot() +
    geom_raster(
        data = hill_multi_tbl,
        aes(x, y, fill = sum),
        show.legend = FALSE
    ) +
    scale_fill_distiller(palette = "Greys") +
    new_scale_fill() +
    geom_raster(
        data = mdt_tbl,
        aes(x, y, fill = alt),
        alpha = .7
    ) +
    scale_fill_hypso_tint_c(
        breaks = c(
            100, 200, 300, 400, 500, 600, 700, 800, 900, 1000
        )
    ) +
    geom_sf(
        data = lakes,
        fill = "#c6dbef",
        color = NA
    ) +
    guides(
        fill = guide_colorsteps(
            barwidth       = .5,
            barheight      = 20,
            title          = "m",
            title.position = "top",
            direction      = "vertical"
        )
    ) +
    coord_sf() +
    theme_void()

p3
```

```{r}
# ggsave(
#     "img/03_finland_multidirectional.png",
#     p3,
#     width  = 8.3,
#     height = 11.7,
#     unit   = "in",
#     device = png,
#     type   = "cairo",
#     bg     = "white"
# )
```

### Using the {ggblend} package

Unlike the previous three, the fourth plot doesn't work properly until it has been saved as a .png file. That's why I'm not even trying to display it here.

```{r}
# map
p4 <- ggplot() +
    list(
        geom_raster(
            data = hill_multi_tbl,
            aes(x, y, fill = sum),
            show.legend = FALSE
        ),
        scale_fill_distiller(palette = "Greys"),
        new_scale_fill(),
        geom_raster(
            data = mdt_tbl,
            aes(x, y, fill = alt),
            alpha = .7
        ),
        scale_fill_hypso_tint_c(
            breaks = c(
                100, 200, 300, 400, 500, 600, 700, 800, 900, 1000
            )
        )
    ) %>% blend("multiply") +
    geom_sf(
        data  = lakes,
        fill  = "#c6dbef",
        color = NA
    ) +
    guides(
        fill = guide_colorsteps(
            barwidth       = .5,
            barheight      = 20,
            title          = "m",
            title.position = "top",
            direction      = "vertical"
        )
    ) +
    coord_sf() +
    theme_void()
```

```{r}
# ggsave(
#     "img/04_finland_multidirectional_ggblend.png",
#     p4,
#     width  = 8.3,
#     height = 11.7,
#     unit   = "in",
#     device = png,
#     type   = "cairo",
#     bg     = "white"
# )
```

## Another alternative for multidirectional shadows

```{r}
# instal whitebox
install_whitebox()
```

```{r}
# export the DEM
writeRaster(
    mdt,
    "img/mdt.tiff",
    overwrite = TRUE)

# launch whitebox
wbt_init()
```

```{r}
# create the hillshade
wbt_multidirectional_hillshade(
    "img/mdt.tiff",
    "img/hillshade.tiff"
)
```

```{r}
# re-import the hillshade
hill_wb <- rast("img/hillshade.tiff")

plot(hill_wb)
```

```{r}
# remask
hill_wb <- mask(
    hill_wb,
    vect(fin)
)

plot(hill_wb)
```

```{r}
# convert the hillshade to xyz
hill_wb_tbl <- as.data.frame(
    hill_wb,
    xy = TRUE
) %>%
    as_tibble()

hill_wb_tbl
```

```{r}
# map
p5 <- ggplot() +
    geom_raster(
        data = hill_wb_tbl,
        aes(x, y, fill = hillshade),
        show.legend = FALSE
    ) +
    scale_fill_distiller(palette = "Greys") +
    new_scale_fill() +
    geom_raster(
        data = mdt_tbl,
        aes(x, y, fill = alt),
        alpha = .7
    ) +
    scale_fill_hypso_tint_c(
        breaks = c(
            100, 200, 300, 400, 500, 600, 700, 800, 900, 1000
        )
    ) +
    geom_sf(
        data  = lakes,
        fill  = "#c6dbef",
        color = NA
    ) +
    guides(
        fill = guide_colorsteps(
            barwidth       = .5,
            barheight      = 20,
            title          = "m",
            title.position = "top",
            direction      = "vertical"
        )
    ) +
    coord_sf() +
    theme_void()

p5
```

```{r}
# ggsave(
#     "img/05_finland_multidirectional_whitebox.png",
#     p5,
#     width  = 8.3,
#     height = 11.7,
#     unit   = "in",
#     device = png,
#     type   = "cairo",
#     bg     = "white"
# )
```
